Determinism, Complexity, and Predictability in Computer Performance

### 2014-04-11 results.tex
adding some notes

The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt
  col\_major}.  This indicates that the rate of forward information
transfer of the underlying process is lower, but that time-series data
from this system still contain a significant amount of structure that
can, in theory, be leveraged to predict the future course of the time
series.  As mentioned above, though, the effectiveness of any
prediction strategy depends on how well it leverages the available
information: methods that use linear models, for instance, may fail
when applied to nonlinear processes.
% 
% (This is the basis for the second conjecture above.)  
% 
This match between model and task is an issue in the \svdfive
experiments as well.  The $MASE$ scores of the \naive ~and ARIMA
predictions for this system are $20.870 \pm 0.192$ and $2.370 \pm
0.051$, respectively: that is, 20.87 and 2.37 times worse than a
simple random walk forecast of the same signals\footnote{The \naive
  ~$MASE$ score is large because of the bimodal nature of the
  distribution of the values of the signal, which makes guessing the
  mean a particularly bad strategy.  The same thing is true of the
  \svdthree signal.}.  As before, the positions of these points on a
WPE vs. $MASE$ plot---significantly below and to the right of the
dashed line---should suggest to a practitioner that there is more
structure in this signal than the corresponding method is able to
leverage, and that a different method might do better.  Indeed, for
\svdfive, the LMA method produces a $MASE$ score of $ 0.718\pm 0.048 $
and a cluster of results that is much closer to the dashed line on the
WPE-$MASE$ plot.  Again, this validates our second conjecture: the LMA
method can capture and reproduce the way in which the \svdfive system
processes information, but the \naive ~and ARIMA prediction methods
cannot.

@@

\svdone behaves very differently than the other seven systems in this
study.  Though its weighted permutation entropy is very high ($0.957
\pm 0.016$), two of the three prediction methods do quite well,
yielding $MASE$ scores of $0.827 \pm 0.076$ (LMA) and $0.714 \pm
0.075$ (ARIMA).  This pushes the corresponding clusters of points in
the bottom two images in Figure~\ref{fig:wpe_vs_mase_all} well above
the trend followed by the other seven clusters.  The $MASE$ scores of
the predictions of this system produced by the \naive ~method,
however, are highly inconsistent: $2.676 \pm 4.328$.  The majority of
the blue diamond-shaped points on the top plot in the figure are
clustered near a $MASE$ score of 0.6, but three of the 15 \svdone
trials posed challenges to this prediction method, resulting in $MASE$
scores $\approx 2.5$.

These effects, we believe, are artifacts of the way $MASE$ is
calculated, as well as the nonstationarity properties of the \svdone
signal.  Recall that $MASE$ scores are scaled \emph{relative to a
  random-walk forecast}, which uses the current value as the
prediction.  This strategy works very badly on signals with frequent,
large, rapid transitions.  Consider a signal that oscillates from one
end of its range to the other at every step.  A signal like this will
have a low WPE, much like {\tt col\_major}.  However, a random-walk
forecast of this signal will be 180 degrees out of phase with the true
continuation.  Since random-walk error appears in the denominator of
the $MASE$ score, this effect can shift points leftwards on a WPE
vs. $MASE$ plot, and that is exactly why the \svdone clusters for
ARIMA and LMA are above the dashed line in
Figure~\ref{fig:wpe_vs_mase_all}.  This time series, which is shown in
closeup in Figure~\ref{fig:svdone-ts},
\begin{figure}[htbp]
  \centering
    \includegraphics[width=\columnwidth]{figs/svdonets2}
\caption{A small portion of the \svdone time series}\label{fig:svdone-ts}
\end{figure} 
is not quite to the level of the worst-case signal described above,
but it still poses a serious challenge to random-walk prediction.  It
is dominated by a white-noise regime (between $\approx$1.86 and
$\approx$1.88 on the vertical scale in the Figure), punctuated by
short excursions above 1.9.  In the former regime, which makes up more
than 80\% of the signal, there are frequent dips to 1.82 and
occasional larger dips below 1.8.  These single-point dips are the
bane of random-walk forecasting.  In this particular case, roughly
40\% of the forecasted points are off by the width of the associated
dip, which skews the associated $MASE$ scores.  Signals like this are
also problematic for the \naive ~prediction strategy, since the
outliers have significant influence on the mean.  This compounds the
effect of the skew in the scaling factor and creates a large spread in
the \svdone $MASE$ values.  {\color{red} We need to think about the
  three outliers and put in a sentence here about them.}  For all of
these reasons, we view \svdone as an outlier and exclude it from the
fit calculation of the dashed line in
Figure~\ref{fig:wpe_vs_mase_best}.

>>>>

The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt
  col\_major}.  This indicates that the rate{\color{red} is this really a rate} of forward information
transfer of the underlying process is lower, but that time-series data
from this system still contain a significant amount of structure that
can, in theory, be leveraged to predict the future course of the time
series.  As mentioned above, though, the effectiveness of any
prediction strategy depends on how well it leverages the available
information: methods that use linear models, for instance, may fail
when applied to nonlinear processes.
% 
% (This is the basis for the second conjecture above.)  
% 
This match between model and task is an issue in the \svdfive
experiments as well.  The $MASE$ scores of the \naive ~and ARIMA
predictions for this system are $20.870 \pm 0.192$ and $2.370 \pm
0.051$, respectively: that is, 20.87 and 2.37 times worse than a
simple random walk forecast of the same signals\footnote{The \naive
  ~$MASE$ score is large because of the bimodal nature of the
  distribution of the values of the signal, which makes guessing the
  mean a particularly bad strategy.  The same thing is true of the
  \svdthree signal.}.  As before, the positions of these points on a
WPE vs. $MASE$ plot---significantly below and to the right of the
dashed line---should suggest to a practitioner that there is more
structure in this signal than the corresponding method is able to
leverage, and that a different method might do better.  Indeed, for
\svdfive, the LMA method produces a $MASE$ score of $ 0.718\pm 0.048 $
and a cluster of results that is much closer to the dashed line on the
WPE-$MASE$ plot.  Again, this validates our second conjecture: the LMA
method can capture and reproduce the way in which the \svdfive system
processes information, but the \naive ~and ARIMA prediction methods
cannot.

@@

\svdone behaves very differently than the other seven systems in this
study.  Though its weighted permutation entropy is very high ($0.957
\pm 0.016$), two of the three prediction methods do quite well,
yielding $MASE$ scores of $0.827 \pm 0.076$ (LMA) and $0.714 \pm
0.075$ (ARIMA).  This pushes the corresponding clusters of points in
the bottom two images in Figure~\ref{fig:wpe_vs_mase_all} well above
the trend followed by the other seven clusters.  The $MASE$ scores of
the predictions of this system produced by the \naive ~method,
however, are highly inconsistent: $2.676 \pm 4.328$.  The majority of
the blue diamond-shaped points on the top plot in the figure are
clustered near a $MASE$ score of 0.6, but three of the 15 \svdone
trials posed challenges to this prediction method, resulting in $MASE$
scores $\approx 2.5$. {\color{red} this is becasue of the weird split it does some of the time. Should this be said?}

These effects, we believe, are artifacts of the way $MASE$ is
calculated, as well as the nonstationarity properties of the \svdone
signal.  Recall that $MASE$ scores are scaled \emph{relative to a
  random-walk forecast}, which uses the current value as the
prediction.  This strategy works very badly on signals with frequent,
large, rapid transitions.  Consider a signal that oscillates from one
end of its range to the other at every step.  A signal like this will
have a low WPE, much like {\tt col\_major}.  However, a random-walk
forecast of this signal will be 180 degrees out of phase with the true
continuation.  Since random-walk error appears in the denominator of
the $MASE$ score, this effect can shift points leftwards on a WPE
vs. $MASE$ plot, and that is exactly why the \svdone clusters for
ARIMA and LMA are above the dashed line in
Figure~\ref{fig:wpe_vs_mase_all}.  This time series, which is shown in
closeup in Figure~\ref{fig:svdone-ts},
\begin{figure}[htbp]
  \centering
    \includegraphics[width=\columnwidth]{figs/svdonets2}
\caption{A small portion of the \svdone time series}\label{fig:svdone-ts}
\end{figure} 
is not quite to the level of the worst-case signal described above,
but it still poses a serious challenge to random-walk prediction.  It
is dominated by a white-noise{\color{red} can we actually say white-noise here} regime (between $\approx$1.86 and
$\approx$1.88 on the vertical scale in the Figure), punctuated by
short excursions above 1.9.  In the former regime, which makes up more
than 80\% of the signal, there are frequent dips to 1.82 and
occasional larger dips below 1.8.  These single-point dips are the
bane of random-walk forecasting.  In this particular case, roughly
40\% of the forecasted points are off by the width of the associated
dip, which skews the associated $MASE$ scores.  Signals like this are
also problematic for the \naive ~prediction strategy, since the
outliers have significant influence on the mean.  This compounds the
effect of the skew in the scaling factor and creates a large spread in
the \svdone $MASE$ values.  {\color{red} We need to think about the
  three outliers and put in a sentence here about them., these are the dip ones}  For all of
these reasons, we view \svdone as an outlier and exclude it from the
fit calculation of the dashed line in
Figure~\ref{fig:wpe_vs_mase_best}.
